### 2017-04-18 paper/paper.tex
paper

\documentclass{article}
\usepackage{spconf,amsmath,graphicx}
%\documentclass[times, 10pt,twocolumn]{article}
%\usepackage{latex8}
\usepackage{times}
% \usepackage{balance}
\usepackage{psfrag,pstricks,pstricks-add,pst-node}
\usepackage{graphicx}
\usepackage{papertweaker}
\usepackage{epsfig}
\usepackage{amsmath,amssymb}
\usepackage{multirow}
\usepackage{caption}
\usepackage{subcaption}
%\usepackage{caption}
%\RequirePackage{longtable}
%\RequirePackage{tabularx}
\RequirePackage{booktabs}

%\documentstyle[times,art10,twocolumn,latex8]{article}

\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\TODO}[1]{\textcolor{red}{TODO #1}}
\renewcommand{\TODO}[1]{~}
\DeclareMathOperator*{\argmax}{argmax}

\newcommand{\vspacetuned}{\vspace{-0.0cm}}
\newcommand{\vspacetunedsub}{\vspace{-0.0cm}}
% \setlength{\textfloatsep}{0.4cm}

\title{A comparative study of batch-construction strategies for recurrent neural networks in MXNet}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% If multiple authors, uncomment and edit the lines shown below.       %%
%% Note that each line must be emphasized {\em } by itself.             %%
%% (by Stephen Martucci, author of spconf.sty).                         %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\makeatletter
%\def\name#1{\gdef\@name{#1\\}}
%\makeatother
%\name{{\em Firstname1 Lastname1, Firstname2 Lastname2, Firstname3 Lastname3,}\\
%      {\em Firstname4 Lastname4, Firstname5 Lastname5, Firstname6 Lastname6,
%      Firstname7 Lastname7}}
%%%%%%%%%%%%%%% End of required multiple authors changes %%%%%%%%%%%%%%%%%

\makeatletter
\def\name#1{\gdef\@name{#1\\}}
\makeatother \name{{\em Patrick Doetsch, Hermann Ney}}

\address{Human Language Technology and Pattern Recognition,
 Computer Science Department, \\
 RWTH Aachen University, 52062 Aachen, Germany \\
  {\small \tt \{doetsch,ney\}@cs.rwth-aachen.de}
}

%\twoauthors{Karen Sp\"{a}rck Jones.}{Department of Speech and Hearing \\
%  Brittania University, Ambridge, Voiceland \\
%  {\small \tt Karen@sh.brittania.edu} }
%  {Rose Tyler}{Department of Linguistics \\
%  University of Speechcity, Speechland \\
%  {\small \tt RTyler@ling.speech.edu} }

%

\begin{document}

  \maketitle
  %
  \begin{abstract}
  	 In this work we compare different methods for sequence selection for mini-batch training
  	 of recurrent neural networks. While popular implementations like Tensorflow and MXNet suggest 
  	 a bucketing approach to improve the parallelization capabilities of the recurrent training process, 
  	 we propose several simple ordering strategies that are able to compete both in training time and 
  	 recognition performance while being conceptually simpler. We compare our method with various other
  	 batch construction strategies on the CHiME-4 noisy speech recognition corpus. The experiments were
  	 done in MXNet and a demo implementation will be provided for free use.
  \end{abstract}
  \noindent{\bf Index Terms}: MXNet, bucketing, batches, recurrent neural networks

  \section{Introduction}
  	Neural network based acoustic modeling became the de-facto standard in automatic speech recognition (ASR)
  	and related tasks. Modeling contextual information over long distances in the input signal hereby showed to 
  	be of fundamental importance for optimal system performance. Modern acoustic models therefore use recurrent 
  	neural network models to encode information over long time gaps. In particular the long short-term memory (LSTM)
  	has been shown to work very well in these tasks and most current state-of-the-art systems incorporate LSTMs
  	in their acoustic models. While it is common practice to optimize a frame-by-frame labeling obtained from a 
  	previously trained system, sequence level criteria that train the acoustic model and the alignment model jointly
  	are becoming increasingly popular. As an example, the connectionist temporal classification (CTC) \cite{CTC}
  	enables fully integrated training of acoustic models without assuming any alignment to be given. Sequence-level
  	criteria however require to train on full utterances, where it is possible to train frame-wise labeled systems
  	on sub-sequences of any resolution.

    Training of recurrent neural networks (RNN) for large vocabulary continuous speech recognition (LVCSR)
    tasks is computationally very expensive and the sequential nature of recurrent process prohibits to
    parallelize the training on an input frame basis. This requires to work on large batches of sequences
    at once and training time as well as recognition performance can vary strongly depending on the choice of how
    batches were put together. The main reason is that combining sequences of different length into the same batch requires 
    to extend the length of each sequence to the longest sequence within the batch, usually by appending zeros. These zero frames 
    are then later on ignored when gradients are computed but they nevertheless require computing power during RNN
    forwarding.

    A straight forward strategy to minimize zero padding is to sort the sequences by length and to partition them into
    batches afterwards. However, there are significant drawbacks to this method: First, the sequence order remains constant
    and therefore the intra-batch variability is very low since the same sequences are usually combined into the same batch. Secondly, the strategy favors to put very similar sequences into a single batch, since sequences of similar length often also share other properties.
    One way to overcome this limitation was proposed within Tensorflow and is also used as recommended 
    behavior in MXNet The idea is to perform a \textit{bucketing} of the training corpus, where each bucket represents
    a range of sequence lengths and each training sample is then assigned to the bucket that corresponds to its length.
    Afterwards a batch is constructed from a randomly chosen bucket by selecting all or some sequences within the it.
    The concept somehow mitigates the issue of zero padding if suitable length ranges can be defined, while still allowing for 
    some kind of randomness at least when sequences are selected within a bucket. However, buckets have to be made very large 
    in order to ensure a sufficiently large variability within batches, On the other hand, making buckets too large will 
    increase training time due to irrelevant computations on zero padded frames. Setting these hyperparameters correctly is
    therefore of fundamental importance for fast and robust acoustic model training.

    In this work we propose a simple batch construction strategy that does not rely on such a hyper-parametrization. The method 
    produces batches with large variability of sequences while at the same time reducing irrelevant computation to a minimum.
    The following sections we are going to give an overview over current batch construction strategies and compare them 
    w.r.t.~training time and variability. We will then derive our proposed method and discuss its properties on a theoretical
    level, followed by an empirical evaluation on the CHiME-4 noisy speech recognition task. All experiments were implemented 
    within the MXNet framework and are available for download on our website.

  \section{Related Work} \label{sec:related}
  	While mini-batch training was studied extensively for feed-forward networks \cite{Li16}, authors rarely reveal the batch construction strategy they used during training when 
  	RNN experiments are reported. This is because the systems are either trained in a frame-wise
  	fashion ~\cite{2016arXiv161005256X} or because the analysis uses sequences very similar length as in \cite{PascanuMB13}. We studied in an earlier work \cite{Doetsch14} how in those cases training on sub-sequences can lead to significant faster and often also more robust training. 
  	In \cite{Laurent16} the problem of having sequences of largely varying length in a batch was identified and 
  	suggested to adapt the proposed batch-normalization method to a frame-level normalization, 
  	although a sequence-level normalization sounds theoretically more reasonable. In \cite{Bengio15} a curriculum learning strategy is proposed where sequences follow a specific scheduling in order to reduce overfitting.

  	Modern machine learning frameworks like Tensorflow \cite{tensorflow} and 
  	MXNet \cite{mxnet} implement a bucketing approach based on the length of the sequence. In  \cite{Khomenko06} the authors extend this idea by selecting optimal sequences within each bucket using a dynamic programming technique.

  \section{Bucketing in MXNet} \label{sec:bucketing}
  Bucketing, borrowed from TensorFlow's sequence training example, is a simple solution. Instead of unrolling the network to the maximum possible sequence, we unroll multiple instances of different lengths (e.g., length 5, 10, 20, 30). During training, we use the most appropriate unrolled model for each mini-batch of data with different sequence length. For RNNs, although those models have different architecture, the parameters are shared in time. So although models in different buckets are selected to train in different mini-batches, essentially the same set of parameters are being optimized. MXNet reuses the internal memory buffers among all executors.

  \section{Proposed Approach} \label{sec:approach}
  At each iteration:
  \begin{enumerate}
  	\item shuffle training data by sequence length
  	\item partition resulting sequence order into $N$ partitions
  	\item sort each bin $n$:
  	\begin{itemize}
  		\item in increasing order if $n$ is odd
  		\item in decreasing order if $n$ is even
  	\end{itemize}
  \end{enumerate}

  \section{Experimental Setup} \label{sec:setup}
  The 4th CHiME Speech Separation and Recognition Challenge
  \cite{Vincent_CSL2016:CHiME4} revisits the CHiME-3 dataset \cite{Barker2015:CHiME3},
  which consists of utterances spoken by speakers in challenging noisy environments recorded using a 6-channel microphone array on a tablet.
  We extracted 16-dimensional MFCC vectors from beam-formed versions of the six sub-corpora in the training data corresponding to each channel
  and used them as features for the neural network models \cite{menne16:chime4System}.
  A CART tying was used to cluster the (roughly 120k possible) allophones into 1500 CART labels.
  We trained a three state HMM without skip transitions in the standard fashion and applied it in a forced alignment mode to the training data
  in order to obtain a frame-wise labeling of the data.

  \section{Experiments} \label{sec:experiments}
   We conducted experiments with different batch construction strategies. The results are reported in \ref{tab:chime:batch}, where the first three rows show results for trivial sequence ordering 
   methods. The last two rows then provide a direct comparison of the bucketing approach as 
   it is implemented in MXNet and the Laplace approach as proposed in this paper.
    %\begin{table}[tbp]
    %\centering
    %\caption{Comparison of runtime and memory requirement for different software packages. The %numbers were averaged over 100 training epochs. 
    %Note that sharp memory usage estimates for Torch and TensorFlow can not be obtained due to %their 5internal memory management.}
    %\label{tab:chime:compare}
    %\begin{tabular}{lrr}
    %  \hline
    %  Toolkit                 & Runtime [sec] & Memory [GB] \\
    %  \hline
    %  RETURNN   &  140     & 0.6     \\
    %  \hline
    %  Theano (in RETURNN)     &  370     & 0.8     \\
    %  TensorFlow              &  580     & 2.1$^*$     \\
    %  \hline
    %\end{tabular}
    %\end{table}

	\begin{table}[tbp]
		\centering
		\caption{Training time and recognition performance when training is done 
		on sub-sequence level. The first column shows the maximal sub-sequence length 
		after partitioning of the original sequence. The last line shows the results 
		if no partitioning into sub-sequence is performed. }
		\label{tab:chime:batch}
		\begin{tabular}{lrrr}
			\hline
			Chunk size              & Runtime [sec] & Memory [GB] & WER [\%] \\
			\hline
			10     					&				&			  &			 \\
			50  					&				&			  & 		 \\
			100 					& 				&			  &			 \\
			500						& 				&			  &			 \\
			$\max$					&				&			  &			 \\
			\hline
		\end{tabular}
	\end{table}

	\begin{table}[tbp]
		\centering
		\caption{An evaluation of different sequence batch construction methods on the CHiME-4 database. Training time per epoch, memory consumption are presented in the first two columns, while the last columns shows the word error rate of the corresponding system
		when decoded with the RASR speech recognition toolkit.}
		\label{tab:chime:batch}
		\begin{tabular}{lrrr}
			\hline
			Approach                & Runtime [sec] & Memory [GB] & WER [\%] \\
			\hline
			Static					&				&			  &			 \\
			Random 					&				&			  & 		 \\
			Sorted					& 				&			  &			 \\
			\hline
			Bucketing				&				&			  &			 \\
			Laplace					&				&			  &			 \\
			\hline
		\end{tabular}
	\end{table}

  As expected, sorting the entire training set by sequence length reduces the required time 
  per epoch to a minimum.

  \section{Conclusions}
    In this work we presented a novel strategy to construct sequence-level batches for recurrent 
    neural network training. Most system rely on a bucketing approach by clustering sequences of similar length into bins and to create batches from each bin individually. We showed that we can achieve a better runtime performance using a simpler batch design.

   \section{Acknowledgements}

   apptek?

   %\newpage
   \ninept
   \bibliographystyle{IEEEtran}
   \bibliography{strings,paper}

\end{document}

### 2017-04-18 code/run.py

@@ -30,15 +30,15 @@ parser.add_argument('--kv-store', type=str, default='device',
                     help='key-value store type')
 parser.add_argument('--num-epochs', type=int, default=25,
                     help='max num of epochs')
-parser.add_argument('--lr', type=float, default=1.0,
+parser.add_argument('--lr', type=float, default=0.1,
                     help='initial learning rate')
-parser.add_argument('--optimizer', type=str, default='adadelta',
+parser.add_argument('--optimizer', type=str, default='sgd',
                     help='the optimizer type')
-parser.add_argument('--mom', type=float, default=0.0,
+parser.add_argument('--mom', type=float, default=0.9,
                     help='momentum for sgd')
 parser.add_argument('--wd', type=float, default=0.00001,
                     help='weight decay for sgd')
-parser.add_argument('--batch-size', type=int, default=5,
+parser.add_argument('--batch-size', type=int, default=25,
                     help='the batch size.')
 parser.add_argument('--disp-batches', type=int, default=100,
                     help='show progress for every n batches')
@@ -93,7 +93,7 @@ class UtteranceIter(DataIter):
       self.labels = [[] for _ in sampling]
       for utt, lab in zip(utterances, states):
         buck = bisect.bisect_left(sampling, len(utt))
-        xin = np.full((sampling[buck],len(utt[0])), 0, dtype='float32')
+        xin = np.full((sampling[buck],len(utt[0])), -1, dtype='float32')
         n_in = len(utt[0])
         xin[:len(utt)] = utt
         yout = np.full((sampling[buck],), -1, dtype='int32')
@@ -150,9 +150,9 @@ class UtteranceIter(DataIter):
       i, j = self.idx[self.curr_idx]
 
       data = self.nddata[i][j:j + self.batch_size]
-      label = self.ndlabel[i][j:j + self.batch_size]
+      label = self.ndlabel[i][j:j + self.batch_size].T
       data = ndarray.swapaxes(data, 1, 0) # TBD
-      label = ndarray.swapaxes(label, 1, 0)
+      #label = ndarray.swapaxes(label, 1, 0)
 
       batch = DataBatch([data], [label], pad=0,
                         bucket_key=self.sampling[i],
@@ -214,11 +214,12 @@ class FrameError(mx.metric.EvalMetric):
     for label, pred_label in zip(labels, preds):
[…]

### 2017-04-18 paper/paper.tex
paper

@@
   \section{Bucketing in MXNet} \label{sec:bucketing}
-  Bucketing, borrowed from TensorFlow's sequence training example, is a simple solution.
-Instead of unrolling the network to the maximum possible sequence, we unroll multiple instances of different lengths (e.g., length 5, 10, 20, 30).
-During training, we use the most appropriate unrolled model for each mini-batch of data with different sequence length.
-For RNNs, although those models have different architecture, the parameters are shared in time.
-So although models in different buckets are selected to train in different mini-batches, essentially the same set of parameters are being optimized.
-MXNet reuses the internal memory buffers among all executors.
+  Borrowed from TensorFlow's sequence training example, MXNet implements bucketing by 
+  clustering sequences into bins depending on their length. The size of each bin, i.e.~ the span of 
+  sequence lengths associated with this bin, has to be specified by the user and optimal values 
+  depend on the ASR task. The sampling process can be done in logarithmic time, since for each sequence length in the training set a binary search over the bins can be performed.
+
+  In each iteration of the mini-batch training a bucket is then selected randomly.
+Within the selected bucket a random span of sequences is chosen to be used as data batch.
+Note that this random shuffling only ensures a large inter-batch variance w.r.t.~the sequence length, while the variance
+  within each batch is small.
+
+  Bucketing is especially suited if the RNN model itself is not able to handle arbitrary 
+  long sequences but instead requires to store an unrolled version of the network for every possible 
+  length. In those cases bucketing allows the framework to assign each batch to the shortest possible 
+  unrolled network, while still optimizing the same shared weights.
 
   \section{Proposed Approach} \label{sec:approach}
+
   At each iteration:
@@
   The 4th CHiME Speech Separation and Recognition Challenge
-  \cite{Vincent_CSL2016:CHiME4} revisits the CHiME-3 dataset \cite{Barker2015:CHiME3},
-  which consists of utterances spoken by speakers in challenging noisy environments recorded using a 6-channel microphone array on a tablet.
+  \cite{Vincent_CSL2016:CHiME4} consists of noisy utterances spoken by speakers in challenging 
+  environments. Recorded was done using a 6-channel microphone array on a tablet. The dataset revisits 
+  the CHiME-3 corpus that was published one year before the challenge took place \cite{Barker2015:CHiME3}.
+
   We extracted 16-dimensional MFCC vectors from beam-formed versions of the six sub-corpora in the training data corresponding to each channel
@@
    methods. The last two rows then provide a direct comparison of the bucketing approach as 
-   it is implemented in MXNet and the Laplace approach as proposed in this paper.
+   it is implemented in MXNet and the sinusoidal approach as proposed in this paper.
     %\begin{table}[tbp]
@@
 			Bucketing				&				&			  &			 \\
-			Laplace					&				&			  &			 \\
+			Sinusoidal  			&				&			  &			 \\
 			\hline

### 2017-04-18 paper/paper.tex
minor edits "intro", bibtex

@@
 
-\title{A comparative study of batch-construction strategies for recurrent neural networks in MXNet}
+\title{A comparative study of batch construction strategies \\
+	 for recurrent neural networks in MXNet}
 
@@
   	 In this work we compare different methods for sequence selection for mini-batch training
-  	 of recurrent neural networks. While popular implementations like Tensorflow and MXNet suggest 
+  	 of recurrent neural networks. While popular implementations like TensorFlow and MXNet suggest 
   	 a bucketing approach to improve the parallelization capabilities of the recurrent training process, 
@@
   	be of fundamental importance for optimal system performance. Modern acoustic models therefore use recurrent 
-  	neural network models to encode information over long time gaps. In particular the long short-term memory (LSTM)
-  	has been shown to work very well in these tasks and most current state-of-the-art systems incorporate LSTMs
-  	in their acoustic models. While it is common practice to optimize a frame-by-frame labeling obtained from a 
-  	previously trained system, sequence level criteria that train the acoustic model and the alignment model jointly
+  	neural networks (RNN) to model long temporal dependencies. In particular the long short-term memory (LSTM)
+  	has been shown to work very well on these tasks and most current state-of-the-art systems incorporate LSTMs
+  	in their acoustic models. While it is common practice to train the
+  	models on a frame-by-frame labeling obtained from a 
+  	previously trained system, sequence-level criteria that optimize the acoustic model and the alignment model jointly
   	are becoming increasingly popular. As an example, the connectionist temporal classification (CTC) \cite{CTC}
-  	enables fully integrated training of acoustic models without assuming any alignment to be given. Sequence-level
+  	enables fully integrated training of acoustic models without assuming a frame-level alignment to be given. Sequence-level
   	criteria however require to train on full utterances, where it is possible to train frame-wise labeled systems
@@
 
-    Training of recurrent neural networks (RNN) for large vocabulary continuous speech recognition (LVCSR)
+    Training of recurrent neural networks for large vocabulary continuous speech recognition (LVCSR)
     tasks is computationally very expensive and the sequential nature of recurrent process prohibits to
-    parallelize the training on an input frame basis. This requires to work on large batches of sequences
-    at once and training time as well as recognition performance can vary strongly depending on the choice of how
+    parallelize the training over input frames. A robust optimization requires to work on large batches of sequences
+    and training time as well as recognition performance can vary strongly depending on the choice of how
     batches were put together. The main reason is that combining sequences of different length into the same batch requires 
     to extend the length of each sequence to the longest sequence within the batch, usually by appending zeros. These zero frames 
-    are then later on ignored when gradients are computed but they nevertheless require computing power during RNN
-    forwarding.
+    are ignored later on when gradients are computed but the forward-propagation of zeros through the RNN is a waste of computing power.
 
-    A straight forward strategy to minimize zero padding is to sort the sequences by length and to partition them into
-    batches afterwards. However, there are significant drawbacks to this method: First, the sequence order remains constant
+    A straight-forward strategy to minimize zero padding is to sort the sequences by length and to partition them into
+    batches afterwards. However, there are significant drawbacks to this method. First, the sequence order remains constant in each epoch 
     and therefore the intra-batch variability is very low since the same sequences are usually combined into the same batch.
-Secondly, the strategy favors to put very similar sequences into a single batch, since sequences of similar length often also share other properties.
-    One way to overcome this limitation was proposed within Tensorflow and is also used as recommended 
+Second, the strategy favors putting very similar sequences into the same batch, since short sequences often tend to share other properties.
+    One way to overcome this limitation was proposed within TensorFlow and is also used as recommended 
     behavior in MXNet The idea is to perform a \textit{bucketing} of the training corpus, where each bucket represents
     a range of sequence lengths and each training sample is then assigned to the bucket that corresponds to its length.
-    Afterwards a batch is constructed from a randomly chosen bucket by selecting all or some sequences within the it.
+    Afterwards a batch is constructed from a randomly chosen bucket by selecting all or some sequences from it.
     The concept somehow mitigates the issue of zero padding if suitable length ranges can be defined, while still allowing for 
-    some kind of randomness at least when sequences are selected within a bucket. However, buckets have to be made very large 
-    in order to ensure a sufficiently large variability within batches, On the other hand, making buckets too large will 
-    increase training time due to irrelevant computations on zero padded frames. Setting these hyperparameters correctly is
+    some level of randomness at least when sequences are selected within a bucket. However, buckets have to be made very large 
+    in order to ensure a sufficiently large variability within batches. On the other hand, making buckets too large will 
+    increase training time due to irrelevant computations on zero padded frames. Setting these hyper-parameters correctly is
     therefore of fundamental importance for fast and robust acoustic model training.

### 2017-04-18 paper/paper.tex
experiments

@@
   A CART tying was used to cluster the (roughly 120k possible) allophones into 1500 CART labels.
-  We trained a three state HMM without skip transitions in the standard fashion and applied it in a forced alignment mode to the training data
-  in order to obtain a frame-wise labeling of the data.
+  We trained a three state HMM without skip transitions in the standard fashion and applied it in a forced alignment mode to the training data in order to obtain a frame-wise labeling of the data.
+A three layer
+  bi-directional LSTM followed a softmax layer was trained to minimize the cross-entropy of the softmax output and the CART label sequence at each frame.
+Optimization was done with the Adam optimizer and a constant learning rate of 0.01.
+
+  After training, state posteriors were extracted from the softmax for each observation of the evaluation set. The posteriors were then decoded in the RWTH 
+  speech recognition toolkit RASR \cite{rybach2011:rasr} using the hybrid HMM approach. A 4-gram LM 
+  was used during decoding with a language model scale of 12.
 
   \section{Experiments} \label{sec:experiments}
-   We conducted experiments with different batch construction strategies. The results are reported in \ref{tab:chime:batch}, where the first three rows show results for trivial sequence ordering 
+   In order to provide some perspective on the required context on the CHiME-4 task, 
+   we provide results for experiments on sub-utterance (chunk) in 
+   Table \ref{tab:chime:chunk}.
+In the first columns we evaluate different sub-sequence lengths, while measuring the utterances that were processed per second, the memory required and the word error rate (WER) on the evaluation set of the CHiME-4 database.
+We can observe that
+   while large speed-ups can be obtained when training is done in this fashion, full-utterance context is
+   required for optimal performance.
+
+   For training on full sequences we conducted experiments with different batch construction strategies.
+   The results are reported in Table \ref{tab:chime:batch}, where the first three rows show results for trivial sequence ordering 
    methods. The last two rows then provide a direct comparison of the bucketing approach as 
-   it is implemented in MXNet and the sinusoidal approach as proposed in this paper.
-    %\begin{table}[tbp]
-    %\centering
-    %\caption{Comparison of runtime and memory requirement for different software packages. The %numbers were averaged over 100 training epochs. 
-    %Note that sharp memory usage estimates for Torch and TensorFlow can not be obtained due to %their 5internal memory management.}
-    %\label{tab:chime:compare}
-    %\begin{tabular}{lrr}
-    %  \hline
-    %  Toolkit                 & Runtime [sec] & Memory [GB] \\
-    %  \hline
-    %  RETURNN   &  140     & 0.6     \\
-    %  \hline
-    %  Theano (in RETURNN)     &  370     & 0.8     \\
-    %  TensorFlow              &  580     & 2.1$^*$     \\
-    %  \hline
-    %\end{tabular}
-    %\end{table}
+   it is implemented in MXNet and the alternated sorting approach as proposed in this paper.
 
@@
 		if no partitioning into sub-sequence is performed. }
-		\label{tab:chime:batch}
+		\label{tab:chime:chunk}
 		\begin{tabular}{lrrr}
 			\hline
-			Chunk size              & Runtime [sec] & Memory [GB] & WER [\%] \\
+			Chunk size              & Utt./sec & Memory [GB] & WER [\%] \\
 			\hline
@@
 			\hline
-			Approach                & Runtime [sec] & Memory [GB] & WER [\%] \\
+			Approach                & Utt./sec & Memory [GB] & WER [\%]          \\
 			\hline
-			Static					&				&			  &			 \\
-			Random 					&				&			  & 		 \\
-			Sorted					& 				&			  &			 \\
+			Random 					&	       & 			 & 6.8		 \\
+			Sorted					& 		   &			 & 7.4		 \\
 			\hline
-			Bucketing				&				&			  &			 \\
-			Sinusoidal  			&				&			  &			 \\
+			Bucketing				&		   &			 & 7.2		 \\
+			Proposed (4 bins)	&		   &			 & 7.0		 \\
+			Proposed (64 bins)	&		   &			 & 7.0		 \\
+			Proposed (256 bins)	&		   &			 & 6.9		 \\
 			\hline
@@
   As expected, sorting the entire training set by sequence length reduces the required time 
-  per epoch to a minimum.
+  per epoch to a minimum, while the best overall performance is obtained when sequences are shuffled randomly.
 
@@
     neural network training.
-Most system rely on a bucketing approach by clustering sequences of similar length into bins and to create batches from each bin individually.
-We showed that we can achieve a better runtime performance using a simpler batch design.
+Most frameworks rely on a bucketing approach by clustering sequences of similar length into bins and to create batches from each bin individually.
+We showed that we can achieve a better runtime performance using a simpler batch design, by partitioning a shuffled sequence order and to sort the partitions in alternating order.
+The method was evaluated on the ChiME-4 noisy speech recognition task and
+    compared to the bucketing approach of MXNet.
 

### 2017-04-19 paper/paper.tex
proposed approach

@@
 
-  At each iteration:
+  In order to improve the intra-batch variability we propose a stochastic bucketing process. At the beginning
+  of each epoch the utterances are arranged randomly and then partitioned into bins of equal size. Each bin
+  is then sorted in alternating directions such that two consecutive bins are sorted in reverse order to each 
+  other. Finally, the constructed ordering is partitioned into batched. The overall algorithm can be summarized as: \\[1ex]
+
+ \begin{minipage}{\textwidth}
+ 	For each epoch
  	\begin{enumerate}
@@
  		\end{itemize}
+ 		\item partition into batches of desired size \\[1ex]
  	\end{enumerate}
+ \end{minipage}
+
+ Due to the initial shuffling and subsequent partitioning the probability for two sequences of any length being 
+ put into the same bin in $\frac{1}{N*(N-1)}$, so by increasing the number of bins, the variability within a 
+ partition decreases quadratically while the variability among different partitions increases. The alternated
+ sorting approach ensures that utterances at the boundaries of two consecutive bins are of similar length such 
+ that the final partitioning into batches requires minimal zero padding.
 
@@
 
-  We extracted 16-dimensional MFCC vectors from beam-formed versions of the six sub-corpora in the training data corresponding to each channel
-  and used them as features for the neural network models \cite{menne16:chime4System}.
-  A CART tying was used to cluster the (roughly 120k possible) allophones into 1500 CART labels.
-  We trained a three state HMM without skip transitions in the standard fashion and applied it in a forced alignment mode to the training data in order to obtain a frame-wise labeling of the data.
-A three layer
+  We extracted 16-dimensional MFCC vectors as in \cite{menne16:chime4System} from the six sub-corpora
+  and used them as features for the neural network models. The allophones of the lexicon were clustered into 
+  1500 state labels using a classification and regression tree (CART).
+  We trained a three state HMM without skip transitions applied it in a forced alignment mode to the training data in order to obtain a frame-wise labeling. A three layer 
   bi-directional LSTM followed a softmax layer was trained to minimize the cross-entropy of the softmax output and the CART label sequence at each frame.

### 2017-04-19 paper/paper.tex
semi-final, 3 WERs missing

@@
   	 recognition performance while being conceptually simpler. We compare our method with various other
-  	 batch construction strategies on the CHiME-4 noisy speech recognition corpus. The experiments were
-  	 done in MXNet and a demo implementation will be provided for free use.
+  	 batch construction strategies on the CHiME-4 noisy speech recognition corpus. 
   \end{abstract}
@@
   	be of fundamental importance for optimal system performance. Modern acoustic models therefore use recurrent 
-  	neural networks (RNN) to model long temporal dependencies. In particular the long short-term memory (LSTM)
+  	neural networks (RNN) to model long temporal dependencies. In particular the long short-term memory (LSTM)~\cite{hochreiter1997lstm}
   	has been shown to work very well on these tasks and most current state-of-the-art systems incorporate LSTMs
@@
     w.r.t.~training time and variability. We will then derive our proposed method and discuss its properties on a theoretical
-    level, followed by an empirical evaluation on the CHiME-4 noisy speech recognition task. All experiments were implemented 
-    within the MXNet framework and are available for download on our website.
+    level, followed by an empirical evaluation on the CHiME-4 noisy speech recognition task. 
+    %All experiments were implemented 
+    %within the MXNet framework and are available for download on our website.
 
@@
 
-  Bucketing is especially suited if the RNN model itself is not able to handle arbitrary 
+  Bucketing is especially useful if the RNN model itself does not support dynamic unrolling and 
+  is not able to handle arbitrary 
   long sequences but instead requires to store an unrolled version of the network for every possible 
@@
   \section{Proposed Approach} \label{sec:approach}
+
+  \begin{figure}
+   	\resizebox{0.45\textwidth}{!}{\includegraphics{plot}}
+   	\caption{Resulting sequence ordering for different batch construction strategies. The graphs 
+   	show the length distribution for 1000 randomly selected samples of the CHiME-4 training set. On the 
+   	vertical axis the length is plotted and on the X axis the utterance order. Bucketing was done with a bucket size of 250 and in the proposed approach we used 12 bins.}
+   	  \label{fig:plot}
+   \end{figure}
 
@@
  		\end{itemize}
- 		\item partition into batches of desired size \\[1ex]
+ 		\item partition resulting order into batches of desired size \\[1ex]
  	\end{enumerate}
@@
  Due to the initial shuffling and subsequent partitioning the probability for two sequences of any length being 
- put into the same bin in $\frac{1}{N*(N-1)}$, so by increasing the number of bins, the variability within a 
+ put into the same bin in $\frac{1}{N\cdot(N-1)}$, so by increasing the number of bins, the variability within a 
  partition decreases quadratically while the variability among different partitions increases. The alternated
@@
  that the final partitioning into batches requires minimal zero padding.
+
+ Figure \ref{fig:plot} shows the length distribution for random and sorted sequence ordering as well as for bucketing in MXNet and the proposed approach.
+Note that in the case of bucketing batches are put
+ together by randomly choosing one of the buckets first, so the ordering does not directly represent 
+ the final set of batches.
 
@@
 In the first columns we evaluate different sub-sequence lengths, while measuring the utterances that were processed per second, the memory required and the word error rate (WER) on the evaluation set of the CHiME-4 database.
-We can observe that
+We hereby constrained batches to only
+   contain 5,000 frames in total, such that the same number of updates is constant in all experiments. We can observe that 
    while large speed-ups can be obtained when training is done in this fashion, full-utterance context is
-   required for optimal performance.
+   required for optimal performance. However, it is worth noting that the memory requirement decreases 
+   significantly when sub-utterance training is applied and especially for unusually long utterances, 
+   sub-utterance training might be the only way to fit the data into GPU memory.
 
@@
 			\hline
-			10     					&				 &			   &			 \\
-			50  					&				 &			   &    		 \\
-			100 					& 				 &			   &			 \\
-			500						& 				 &			   &			 \\
-			$\max$					&				 &			   &			 \\
+			10     					&	36.7   & 1.6         &			 \\
+			50  					&	31.1   & 1.6		 &    		 \\
+			100 					& 	29.6   & 1.6		 &			 \\
+			500						& 	17.3   & 1.6         &			 \\
+			$\max$					&	7.0	   & 5.4         &	8.9		 \\
 			\hline
@@
 			\hline
-			Random 					&	       & 			 & 6.8		 \\
-			Sorted					& 		   &			 & 7.4		 \\
+			Random 					& $\sim$ 7.0     & $\sim$ 5.4		 & $\sim$ 8.9		 \\
+			Sorted					& 10.2    &	4.2 & 10.2		 \\
 			\hline
-			Bucketing				&		   &			 & 7.2		 \\
-			Proposed (4 bins)	    &		   &			 & 7.0		 \\
-			Proposed (64 bins)	    &		   &			 & 7.0		 \\
-			Proposed (256 bins)	    &		   &			 & 6.9		 \\
+			Bucketing				& 9.5     &	6.3		 & 9.6		 \\
+			Proposed (8 bins)	    & 10.1	   & 4.8     & 9.9		 \\
+			Proposed (64 bins)	    & 10.0     & 5.3     & 9.5		 \\
+			Proposed (256 bins)	    & 8.8	   & 6.0     & 9.1		 \\
 			\hline
@@
   As expected, sorting the entire training set by sequence length reduces the required time 
-  per epoch to a minimum, while the best overall performance is obtained when sequences are shuffled randomly.
+  per epoch to a minimum, while the best overall performance is obtained when sequences are shuffled randomly. Both bucketing and the proposed 
+  approach are in between. We can observe that our method is able to reach almost the same recognition performance as using a randomly shuffled sequence
+  ordering, while being almost as fast as the sorted sequence scheduler. This allows for a good trade-off between runtime and system performance.
 
@@
 
-   apptek?
+   This work has received funding from the European Research Council
+   (ERC) under the European Union’s Horizon 2020 research and innovation
+   program (SEQCLAS, grant agreement No 694537).  The work reflects
+   only the authors' views and the European Research Council Executive
+   Agency is not responsible for any use that may be made of the
+   information it contains.  We further want to thank Jahn Heymann, Lukas
+   Drude and Reinhold H\"ab-Umbach from University of Paderborn, Germany
+   for their CHiME-4 front-end which we used in this work.
 

### 2017-04-19 code/run.py

@@ -21,7 +21,7 @@ parser.add_argument('--num-layers', type=int, default=3,
                     help='number of stacked RNN layers')
 parser.add_argument('--num-hidden', type=int, default=512,
                     help='hidden layer size')
-parser.add_argument('--bidirectional', type=bool, default=True,
+parser.add_argument('--bidirectional', type=bool, default=False,
                     help='whether to use bidirectional layers')
 parser.add_argument('--gpus', type=str,
                     help='list of gpus to run, e.g. 0 or 0,2,5. empty means using cpu. ' \
@@ -30,15 +30,15 @@ parser.add_argument('--kv-store', type=str, default='device',
                     help='key-value store type')
 parser.add_argument('--num-epochs', type=int, default=25,
                     help='max num of epochs')
-parser.add_argument('--lr', type=float, default=0.1,
+parser.add_argument('--lr', type=float, default=0.01,
                     help='initial learning rate')
-parser.add_argument('--optimizer', type=str, default='sgd',
+parser.add_argument('--optimizer', type=str, default='adam',
                     help='the optimizer type')
 parser.add_argument('--mom', type=float, default=0.9,
                     help='momentum for sgd')
 parser.add_argument('--wd', type=float, default=0.00001,
                     help='weight decay for sgd')
-parser.add_argument('--batch-size', type=int, default=25,
+parser.add_argument('--batch-size', type=int, default=10,
                     help='the batch size.')
 parser.add_argument('--disp-batches', type=int, default=100,
                     help='show progress for every n batches')
@@ -80,7 +80,11 @@ class UtteranceIter(DataIter):
   def __init__(self, utterances, states, names, batch_size, sampling, data_name='data', label_name='labels', shuffle=True):
     super(UtteranceIter, self).__init__()
     if not sampling:
-      sampling = [i for i, j in enumerate([len(x) for x in utterances])] #[500::500]
+      minpad = 100
+      #sampling = [i for i, j in enumerate(np.bincount([len(s) for s in utterances]))]#[j for i, j in enumerate(set([len(x) for x in utterances]))] #[500::500]
+      sampling = range(minpad,max([len(s) for s in utterances]),minpad)
+    print sampling
+    #assert False
 
     self.idx = []
     if isinstance(sampling, list):
[…]

### 2017-04-19 plot/plot.py

@@ -1,0 +1,89 @@
+import sys
+import bisect
+import numpy as np
+import matplotlib
+import matplotlib.pyplot as plt
+import matplotlib.patches as mpatches
+
+
+font = {'family': 'serif',
+        'color':  'black',
+        'weight': 'normal',
+        'size': 25,
+        }
+matplotlib.rcParams.update({'font.size': 22})
+
+#sampling = 40
+y = [ int(line.strip()) for line in open(sys.argv[1]).readlines() ][:1000]
+x = range(1,len(y)+1)
+
+x = np.asarray(x)
+
+plt.subplot(411) # random
+plt.xlim(-10,1010)
+plt.text(3, 2200, 'Random', fontdict=font)
+plt.ylabel('Utterance length')
+plt.xlabel('Utterance order')
+yr = np.asarray(y)
+
+sc = plt.scatter(x,yr, label='random')
+
+
+plt.subplot(412) # sorted
+plt.xlim(-10,1010)
+plt.text(3, 2200, 'Sorted', fontdict=font)
+
+ys = sorted(y)
+ys = np.asarray(ys)
+
+plt.scatter(x,ys)
+plt.ylabel('Utterance length')
+plt.xlabel('Utterance order')
+
+plt.subplot(413) # bucket
+plt.xlim(-10,1010)
+plt.text(3, 2200, 'Bucketing', fontdict=font)
+
+buckets = [0] + range(250,max(y),250) + [max(y)]
+bucket_map = {}
+for l in y:
+  b = bisect.bisect_left(buckets, l)
+  if not buckets[b] in bucket_map:
+    bucket_map[buckets[b]] = []
+  bucket_map[buckets[b]].append(l)
+  
+yb = []
+for k in buckets:
+  if k in bucket_map.keys():
+    yb += bucket_map[k]
+
+plt.scatter(x,yb)
+plt.ylabel('Utterance length')
+plt.xlabel('Utterance order')
+
+plt.subplot(414) # laplace
+plt.xlim(-10,1010)
+plt.text(3, 2200, 'Proposed approach', fontdict=font)
+
+N = 12
+start = 0
+end = len(y) / N
+yl = []
+i = 0
+while start < len(y):
+  i += 1
+  if i % 2 == 1:
+    yl += sorted(y[start:end])
+  else:
+    yl += sorted(y[start:end])[::-1]
+  start = end
+  end += len(y) / N
+  end = min(len(y),end)
+
[…]

### 2017-04-19 paper/paper.tex
semi-final, 3 WERs missing

@@
     One way to overcome this limitation was proposed within TensorFlow and is also used as recommended 
-    behavior in MXNet The idea is to perform a \textit{bucketing} of the training corpus, where each bucket represents
+    behavior in MXNet. The idea is to perform a \textit{bucketing} of the training corpus, where each bucket represents
     a range of sequence lengths and each training sample is then assigned to the bucket that corresponds to its length.
@@
 Optimization was done with the Adam optimizer and a constant learning rate of 0.01.
+We used MXNet \cite{mxnet} for experiments using the bucketing approach and RETURNN \cite{doetsch2017:returnn} for the other methods.
 
@@
     In this work we presented a novel strategy to construct sequence-level batches for recurrent 
-    neural network training.
-Most frameworks rely on a bucketing approach by clustering sequences of similar length into bins and to create batches from each bin individually.
+    neural network acoustic model training. While not much attention is given to the topic 
+    of batch construction, we demonstrate that different strategies can lead to large variations 
+    both in training time and recognition performance.
+Most deep-learning frameworks rely on a bucketing approach by clustering sequences of similar length into bins and to create batches from each bin individually.
 We showed that we can achieve a better runtime performance using a simpler batch design, by partitioning a shuffled sequence order and to sort the partitions in alternating order.
 The method was evaluated on the ChiME-4 noisy speech recognition task and
-    compared to the bucketing approach of MXNet.
+    compared to standard approaches like random sequence shuffling and the bucketing approach of MXNet,
+    where our method was able to reach a better trade-off between training time and recognition performance while being easier to parametrize than the bucketing method.
 

### 2017-04-19 paper/paper.tex
proof reading, draft watermark

@@
 \usepackage{subcaption}
+\usepackage[printwatermark]{xwatermark}
+\newwatermark[allpages,color=red!50,angle=45,scale=3,xpos=0,ypos=0]{DRAFT}
+
 %\usepackage{caption}
@@
   \begin{abstract}
-  	 In this work we compare different methods for sequence selection for mini-batch training
+  	 In this work we compare different batch construction methods for mini-batch training
   	 of recurrent neural networks. While popular implementations like TensorFlow and MXNet suggest 
   	 a bucketing approach to improve the parallelization capabilities of the recurrent training process, 
-  	 we propose several simple ordering strategies that are able to compete both in training time and 
-  	 recognition performance while being conceptually simpler. We compare our method with various other
-  	 batch construction strategies on the CHiME-4 noisy speech recognition corpus. 
+  	 we propose a simple ordering strategy that arranges in a stochastic alternating sorted way. We compare our method to sequence bucketing as well as various other
+  	 batch construction strategies on the CHiME-4 noisy speech recognition corpus. The experiments
+  	 show that our alternated sorting approach is able to compete both in training time and recognition performance while being conceptually simpler to implement. 
   \end{abstract}
-  \noindent{\bf Index Terms}: MXNet, bucketing, batches, recurrent neural networks
+  \noindent{\bf Index Terms}: bucketing, batches, recurrent neural networks
 
@@
   	has been shown to work very well on these tasks and most current state-of-the-art systems incorporate LSTMs
-  	in their acoustic models. While it is common practice to train the
+  	into their acoustic models. While it is common practice to train the
   	models on a frame-by-frame labeling obtained from a 
@@
   	criteria however require to train on full utterances, where it is possible to train frame-wise labeled systems
-  	on sub-sequences of any resolution.
+  	on sub-utterances of any resolution.
 
@@
     tasks is computationally very expensive and the sequential nature of recurrent process prohibits to
-    parallelize the training over input frames. A robust optimization requires to work on large batches of sequences
+    parallelize the training over input frames. A robust optimization requires to work on large batches of utterances
     and training time as well as recognition performance can vary strongly depending on the choice of how
-    batches were put together. The main reason is that combining sequences of different length into the same batch requires 
-    to extend the length of each sequence to the longest sequence within the batch, usually by appending zeros. These zero frames 
+    batches were put together. The main reason is that combining utterances of different length into the same batch requires 
+    to extend the length of each utterance to the longest utterance within the batch, usually by appending zeros. These zero frames 
     are ignored later on when gradients are computed but the forward-propagation of zeros through the RNN is a waste of computing power.
 
-    A straight-forward strategy to minimize zero padding is to sort the sequences by length and to partition them into
+    A straight-forward strategy to minimize zero padding is to sort the utterances by length and to partition them into
     batches afterwards. However, there are significant drawbacks to this method. First, the sequence order remains constant in each epoch 
     and therefore the intra-batch variability is very low since the same sequences are usually combined into the same batch.
-Second, the strategy favors putting very similar sequences into the same batch, since short sequences often tend to share other properties.
+Second, the strategy favors putting similar utterances into the same batch, since short utterances often tend to share other properties.
     One way to overcome this limitation was proposed within TensorFlow and is also used as recommended 
-    behavior in MXNet. The idea is to perform a \textit{bucketing} of the training corpus, where each bucket represents
-    a range of sequence lengths and each training sample is then assigned to the bucket that corresponds to its length.
+    strategy in MXNet. The idea is to perform a \textit{bucketing} of the training corpus, where each bucket represents
+    a range of utterance lengths and each training sample is assigned to the bucket that corresponds to its length.
     Afterwards a batch is constructed from a randomly chosen bucket by selecting all or some sequences from it.
@@
 
-    In this work we propose a simple batch construction strategy that does not rely on such a hyper-parametrization. The method 
+    In this work we propose a simple batch construction strategy that is easier to parametrize and
+    implement. The method 
     produces batches with large variability of sequences while at the same time reducing irrelevant computation to a minimum.
@@
   	RNN experiments are reported. This is because the systems are either trained in a frame-wise
-  	fashion ~\cite{2016arXiv161005256X} or because the analysis uses sequences very similar length as in \cite{PascanuMB13}.
+  	fashion ~\cite{2016arXiv161005256X} or because the analysis uses sequences of very similar length as in \cite{PascanuMB13}.
 We studied in an earlier work \cite{Doetsch14} how in those cases training on sub-sequences can lead to significant faster and often also more robust training.
-  	In \cite{Laurent16} the problem of having sequences of largely varying length in a batch was identified and 
-  	suggested to adapt the proposed batch-normalization method to a frame-level normalization, 
+  	In \cite{Laurent16} the problem of having sequences of largely varying lengths in a batch was identified and the authors
+  	suggested to adapt their proposed batch-normalization method to a frame-level normalization, 
   	although a sequence-level normalization sounds theoretically more reasonable.
@@
 
-  	Modern machine learning frameworks like Tensorflow \cite{tensorflow} and 
-  	MXNet \cite{mxnet} implement a bucketing approach based on the length of the sequence.
+  	Modern machine learning frameworks like TensorFlow \cite{tensorflow} and 
+  	MXNet \cite{mxnet} implement a bucketing approach based on the lengths distribution of the sequences.
 In \cite{Khomenko06} the authors extend this idea by selecting optimal sequences within each bucket using a dynamic programming technique.
@@
   sequence lengths associated with this bin, has to be specified by the user and optimal values 
-  depend on the ASR task. The sampling process can be done in logarithmic time, since for each sequence length in the training set a binary search over the bins can be performed.
+  depend on the ASR task. The sampling process can be done in logarithmic time, since for each sequence length in the training set a binary search over the bins has to be performed.
 
@@
   \cite{Vincent_CSL2016:CHiME4} consists of noisy utterances spoken by speakers in challenging 
-  environments. Recorded was done using a 6-channel microphone array on a tablet. The dataset revisits 
+  environments. Recording was done using a 6-channel microphone array on a tablet. The dataset revisits 
   the CHiME-3 corpus that was published one year before the challenge took place \cite{Barker2015:CHiME3}.
@@
   We extracted 16-dimensional MFCC vectors as in \cite{menne16:chime4System} from the six sub-corpora
-  and used them as features for the neural network models. The allophones of the lexicon were clustered into 
+  and used them as features for the neural network models. The context-dependent phonemes extracted from the lexicon were clustered into 
   1500 state labels using a classification and regression tree (CART).
@@
    In order to provide some perspective on the required context on the CHiME-4 task, 
-   we provide results for experiments on sub-utterance (chunk) in 
+   we performed experiments on sub-utterance (chunk) which are presented in 
    Table \ref{tab:chime:chunk}.
-In the first columns we evaluate different sub-sequence lengths, while measuring the utterances that were processed per second, the memory required and the word error rate (WER) on the evaluation set of the CHiME-4 database.
+In the first columns we evaluate different sub-utterance lengths, while measuring the utterances that were processed per second, the memory required and the word error rate (WER) on the evaluation set of the CHiME-4 database.
 We hereby constrained batches to only
@@
    The results are reported in Table \ref{tab:chime:batch}, where the first three rows show results for trivial sequence ordering 
-   methods. The last two rows then provide a direct comparison of the bucketing approach as 
+   methods and the last two rows provide a direct comparison of the bucketing approach as 
    it is implemented in MXNet and the alternated sorting approach as proposed in this paper.
@@
 
-  As expected, sorting the entire training set by sequence length reduces the required time 
-  per epoch to a minimum, while the best overall performance is obtained when sequences are shuffled randomly. Both bucketing and the proposed 
+  As expected, sorting the entire training set by utterance length reduces the required time 
+  per epoch to a minimum, while the best overall performance is obtained when utterances are shuffled randomly. Both bucketing and the proposed 
   approach are in between. We can observe that our method is able to reach almost the same recognition performance as using a randomly shuffled sequence
-  ordering, while being almost as fast as the sorted sequence scheduler. This allows for a good trade-off between runtime and system performance.
+  ordering, while being almost as fast as the sorted utterance scheduler. This allows for a good trade-off between runtime and system performance.
 

### 2017-04-19 code/run.py
correct shuffling

@@ -30,7 +30,7 @@ parser.add_argument('--kv-store', type=str, default='device',
                     help='key-value store type')
 parser.add_argument('--num-epochs', type=int, default=25,
                     help='max num of epochs')
-parser.add_argument('--lr', type=float, default=0.01,
+parser.add_argument('--lr', type=float, default=0.0001,
                     help='initial learning rate')
 parser.add_argument('--optimizer', type=str, default='adam',
                     help='the optimizer type')
@@ -80,7 +80,7 @@ class UtteranceIter(DataIter):
   def __init__(self, utterances, states, names, batch_size, sampling, data_name='data', label_name='labels', shuffle=True):
     super(UtteranceIter, self).__init__()
     if not sampling:
-      minpad = 100
+      minpad = 64
       #sampling = [i for i, j in enumerate(np.bincount([len(s) for s in utterances]))]#[j for i, j in enumerate(set([len(x) for x in utterances]))] #[500::500]
       sampling = range(minpad,max([len(s) for s in utterances]),minpad)
     print sampling
@@ -100,7 +100,7 @@ class UtteranceIter(DataIter):
         xin = np.full((sampling[buck],len(utt[0])), -1, dtype='float32')
         n_in = len(utt[0])
         xin[:len(utt)] = utt
-        yout = np.full((sampling[buck],), -1, dtype='int32')
+        yout = np.full((sampling[buck],), -1, dtype='float32')
         yout[:len(utt)] = lab
         self.data[buck].append(xin)
         self.labels[buck].append(yout)
@@ -111,7 +111,7 @@ class UtteranceIter(DataIter):
       raise NotImplementedError('sampling %s not supported' % str(sampling))
 
     self.data = [np.asarray(i, dtype='float32') for i in self.data] # BTD
-    self.labels = [np.asarray(i, dtype='int32') for i in self.labels] # BT
+    self.labels = [np.asarray(i, dtype='float32') for i in self.labels] # BT
     self.curr_idx = 0
 
     self.batch_size = batch_size
@@ -137,8 +137,11 @@ class UtteranceIter(DataIter):
     if isinstance(self.sampling, list):
       if self.shuffle:
         random.shuffle(self.idx) # shuffle bucket index
-        for buck in self.data: # shuffle sequence index within bucket
-          np.random.shuffle(buck)
+        for buck_utt, buck_lab in zip(self.data,self.labels): # shuffle sequence index within bucket
[…]

### 2017-04-19 paper/paper.tex
title

@@
 		\caption{Training time and recognition performance when training is done 
-		on sub-sequence level. The first column shows the maximal sub-sequence length 
+		on sub-utterances level. The first column shows the maximal sub-utterance length 
 		after partitioning of the original sequence. The last line shows the results 
-		if no partitioning into sub-sequence is performed. }
+		if no partitioning into sub-utterances is performed. }
 		\label{tab:chime:chunk}
@@
 			\hline
-			10     					&	36.7   & 1.6         &			 \\
-			50  					&	31.1   & 1.6		 &    		 \\
-			100 					& 	29.6   & 1.6		 &			 \\
-			500						& 	17.3   & 1.6         &			 \\
+			10     					&	36.7   & 1.6         &	21.3     \\
+			50  					&	31.1   & 1.6		 &  10.1     \\
+			100 					& 	29.6   & 1.6		 &	9.2      \\
+			500						& 	17.3   & 1.6         &	9.0      \\
 			$\max$					&	7.0	   & 5.4         &	8.9      \\
@@
 			\hline
-			Bucketing				& 9.5     &	6.3		 & 9.6		 \\
+			Bucketing (MXNet)		& 9.5     &	6.3		 & 9.6		 \\
 			Proposed (8 bins)	    & 10.1	   & 4.8     & 9.9		 \\

### 2017-04-19 paper/paper.tex
some updates

@@
   	 a bucketing approach to improve the parallelization capabilities of the recurrent training process, 
-  	 we propose a simple ordering strategy that arranges in a stochastic alternating sorted way. We compare our method to sequence bucketing as well as various other
+  	 we propose a simple ordering strategy that arranges the training sequences in a stochastic alternatingly sorted way. We compare our method to sequence bucketing as well as various other
   	 batch construction strategies on the CHiME-4 noisy speech recognition corpus. The experiments
@@
   	enables fully integrated training of acoustic models without assuming a frame-level alignment to be given. Sequence-level
-  	criteria however require to train on full utterances, where it is possible to train frame-wise labeled systems
+  	criteria however require to train on full utterances, while it is possible to train frame-wise labeled systems
   	on sub-utterances of any resolution.
@@
     and training time as well as recognition performance can vary strongly depending on the choice of how
-    batches were put together. The main reason is that combining utterances of different length into the same batch requires 
-    to extend the length of each utterance to the longest utterance within the batch, usually by appending zeros. These zero frames 
+    batches were put together. The main reason is that combining utterances of different lengths in a mini-batch requires 
+    to extend the length of each utterance to that of the longest utterance within the batch, usually by appending zeros. These zero frames 
     are ignored later on when gradients are computed but the forward-propagation of zeros through the RNN is a waste of computing power.
@@
     a range of utterance lengths and each training sample is assigned to the bucket that corresponds to its length.
-    Afterwards a batch is constructed from a randomly chosen bucket by selecting all or some sequences from it.
+    Afterwards a batch is constructed by drawing sequences from a randomly chosen bucket.
     The concept somehow mitigates the issue of zero padding if suitable length ranges can be defined, while still allowing for 
@@
     produces batches with large variability of sequences while at the same time reducing irrelevant computation to a minimum.
-    The following sections we are going to give an overview over current batch construction strategies and compare them 
+    In the following sections we are going to give an overview over current batch construction strategies and compare them 
     w.r.t.~training time and variability. We will then derive our proposed method and discuss its properties on a theoretical
@@
   	RNN experiments are reported. This is because the systems are either trained in a frame-wise
-  	fashion ~\cite{2016arXiv161005256X} or because the analysis uses sequences of very similar length as in \cite{PascanuMB13}.
-We studied in an earlier work \cite{Doetsch14} how in those cases training on sub-sequences can lead to significant faster and often also more robust training.
+  	fashion~\cite{2016arXiv161005256X} or because the analysis uses sequences of very similar length as in \cite{PascanuMB13}.
+We studied in an earlier work \cite{Doetsch14} how training on sub-sequences in those cases can lead to significantly faster and often also more robust training.
   	In \cite{Laurent16} the problem of having sequences of largely varying lengths in a batch was identified and the authors
@@
   Borrowed from TensorFlow's sequence training example, MXNet implements bucketing by 
-  clustering sequences into bins depending on their length. The size of each bin, i.e.~ the span of 
+  clustering sequences into bins depending on their length. The size of each bin, i.e.~the span of 
   sequence lengths associated with this bin, has to be specified by the user and optimal values 
@@
 Note that this random shuffling only ensures a large inter-batch variance w.r.t.~the sequence length, while the variance
-  within each batch is small.
+  within each batch can be small.
 
@@
   \begin{figure}
-   	\resizebox{0.45\textwidth}{!}{\includegraphics{plot}}
+   	\resizebox{0.48\textwidth}{!}{\includegraphics{plot}}
    	\caption{Resulting sequence ordering for different batch construction strategies. The graphs 
-   	show the length distribution for 1000 randomly selected samples of the CHiME-4 training set. On the 
-   	vertical axis the length is plotted and on the X axis the utterance order. Bucketing was done with a bucket size of 250 and in the proposed approach we used 12 bins.}
+   	show the utternace lengths of 1000 randomly selected samples of the CHiME-4 training set. The 
+   	Y-axis shows the length and the X-axis the utterance index for a given ordering. Bucketing was done with a bucket size of 250 and in the proposed approach we used 12 bins.}
    	  \label{fig:plot}
+   	  \vspace{-5mm}
    \end{figure}
@@
   is then sorted in alternating directions such that two consecutive bins are sorted in reverse order to each 
-  other. Finally, the constructed ordering is partitioned into batched. The overall algorithm can be summarized as: \\[1ex]
+  other. Finally, the constructed ordering is partitioned into batches. The overall algorithm can be summarized as follows: \\[1ex]
 
  \begin{minipage}{\textwidth}
+ 	\vspace{-4mm}
  	For each epoch
  	\begin{enumerate}
- 		\item shuffle training data by sequence length
- 		\item partition resulting sequence order into $N$ partitions
- 		\item sort each bin $n$:
+ 		\item shuffle training data % by sequence length
+ 		\item partition resulting sequence into $N$ bins
+ 		\item sort each bin $n$ by the utterance length:
  		\begin{itemize}
- 			\item in increasing order if $n$ is odd
- 			\item in decreasing order if $n$ is even
+ 			\item in ascending order if $n$ is odd
+ 			\item in descending order if $n$ is even
  		\end{itemize}
- 		\item partition resulting order into batches of desired size \\[1ex]
+ 		\item draw consecutive batches of desired size \\
+ 		from the resulting sequence \\[1ex]
  	\end{enumerate}
@@
  Due to the initial shuffling and subsequent partitioning the probability for two sequences of any length being 
- put into the same bin in $\frac{1}{N\cdot(N-1)}$, so by increasing the number of bins, the variability within a 
+ put into the same bin is $\frac{1}{N\cdot(N-1)}$, so by increasing the number of bins, the variability within a 
  partition decreases quadratically while the variability among different partitions increases. The alternated
@@
 
- Figure \ref{fig:plot} shows the length distribution for random and sorted sequence ordering as well as for bucketing in MXNet and the proposed approach.
+ Figure \ref{fig:plot} shows the utterance lengths for random and sorted sequence ordering as well as for bucketing in MXNet and the proposed approach.
 Note that in the case of bucketing batches are put

### 2017-04-19 paper/paper.tex
final pass

@@
   We extracted 16-dimensional MFCC vectors as in \cite{menne16:chime4System} from the six sub-corpora
-  and used them as features for the neural network models. The context-dependent phonemes extracted from the lexicon were clustered into 
-  1500 state labels using a classification and regression tree (CART).
-  We trained a three state HMM without skip transitions applied it in a forced alignment mode to the training data in order to obtain a frame-wise labeling. A three layer 
-  bi-directional LSTM followed a softmax layer was trained to minimize the cross-entropy of the softmax output and the CART label sequence at each frame.
+  and used them as features for the neural network models. The context-dependent HMM states were clustered into 
+  1500 classes using a classification and regression tree (CART).
+  We trained a GMM-based baseline model with three state HMM without skip transitions in order to obtain a frame-wise labeling of the training data. A network of three 
+  bi-directional LSTM layers followed by a softmax layer was trained to minimize the frame-wise cross-entropy.
 Optimization was done with the Adam optimizer and a constant learning rate of 0.01.
@@
 
-  After training, state posteriors were extracted from the softmax for each observation of the evaluation set. The posteriors were then decoded in the RWTH 
-  speech recognition toolkit RASR \cite{rybach2011:rasr} using the hybrid HMM approach. A 4-gram LM 
+  After training, the state posterior estimates from the neural network are normalized by the state priors and used as likelihoods in a conventional hybrid HMM decoder using the RASR toolkit~\cite{rybach2011:rasr}.
+  A 4-gram LM 
   was used during decoding with a language model scale of 12.
@@
   \section{Experiments} \label{sec:experiments}
-   In order to provide some perspective on the required context on the CHiME-4 task, 
+   In order to provide some perspective on the impact of temporal context on the CHiME-4 task, 
    we performed experiments on sub-utterance (chunk) which are presented in 
-   Table \ref{tab:chime:chunk}.
-In the first columns we evaluate different sub-utterance lengths, while measuring the utterances that were processed per second, the memory required and the word error rate (WER) on the evaluation set of the CHiME-4 database.
-We hereby constrained batches to only
-   contain 5,000 frames in total, such that the same number of updates is constant in all experiments. We can observe that 
+   Table~\ref{tab:chime:chunk}.
+For different sub-utterance lengths, we report the processing speed measured in utterances per second, the memory required and the word error rate (WER) on the evaluation set of the CHiME-4 database.
+Here we constrained batches to only
+   contain 5,000 frames in total, such that the overall number of updates is constant in all experiments. We can observe that 
    while large speed-ups can be obtained when training is done in this fashion, full-utterance context is
    required for optimal performance. However, it is worth noting that the memory requirement decreases 
-   significantly when sub-utterance training is applied and especially for unusually long utterances, 
+   significantly when sub-utterance training is applied. In particular, for unusually long utterances, 
    sub-utterance training might be the only way to fit the data into GPU memory.
@@
    For training on full sequences we conducted experiments with different batch construction strategies.
-   The results are reported in Table \ref{tab:chime:batch}, where the first three rows show results for trivial sequence ordering 
-   methods and the last two rows provide a direct comparison of the bucketing approach as 
+   The results are reported in Table \ref{tab:chime:batch}, where the first two rows show results for trivial sequence ordering 
+   methods and the last rows provide a direct comparison of the bucketing approach as 
    it is implemented in MXNet and the alternated sorting approach as proposed in this paper.
@@
 		\caption{Training time and recognition performance when training is done 
-		on sub-utterances level. The first column shows the maximal sub-utterance length 
-		after partitioning of the original sequence. The last line shows the results 
-		if no partitioning into sub-utterances is performed. }
+		on sub-utterance level. The first column shows the maximal sub-utterance length 
+		after partitioning of the original sequence. The last row shows the results 
+		obtained without partitioning into sub-utterances. }
 		\label{tab:chime:chunk}
@@
 		\caption{An evaluation of different sequence batch construction methods on the CHiME-4 database.
-Training time per epoch, memory consumption are presented in the first two columns, while the last columns shows the word error rate of the corresponding system
-		when decoded with the RASR speech recognition toolkit.}
+Training time per epoch, memory consumption are presented in the first two columns, while the last column shows the word error rate of the corresponding acoustic model on the evaluation set.}
 		\label{tab:chime:batch}
-		\begin{tabular}{lrrr}
+		\begin{tabular}{l@{} rrr}
 			\hline
@@
     both in training time and recognition performance.
-Most deep-learning frameworks rely on a bucketing approach by clustering sequences of similar length into bins and to create batches from each bin individually.
-We showed that we can achieve a better runtime performance using a simpler batch design, by partitioning a shuffled sequence order and to sort the partitions in alternating order.
+Most deep-learning frameworks rely on a bucketing approach by clustering sequences of similar length into bins and to draw batches from each bin individually.
+We showed that we can achieve a better runtime performance using a simpler batch design, by partitioning a shuffled sequence order and to sort the partitions in an alternating order.
 The method was evaluated on the ChiME-4 noisy speech recognition task and

### 2017-04-19 paper/paper.tex
add Pavel

@@
 \def\name#1{\gdef\@name{#1\\}}
-\makeatother \name{{\em Patrick Doetsch, Hermann Ney}}
+\makeatother \name{{\em Patrick Doetsch, Pavel Golik, Hermann Ney}}
 
@@
  RWTH Aachen University, 52062 Aachen, Germany \\
-  {\small \tt \{doetsch,ney\}@cs.rwth-aachen.de}
+  {\small \tt \{doetsch,golik,ney\}@cs.rwth-aachen.de}
 }

### 2017-04-20 code/run.py
more flexible forwarding

@@ -17,6 +17,8 @@ parser.add_argument('--load-epoch', type=int, default=0,
                     help='load from epoch')
 parser.add_argument('--output-file', type=str, default='out.cache',
                     help='posterior output cache file')
+parser.add_argument('--input-file', type=str, default='data/valid.0001',
+                    help='feature input file')
 parser.add_argument('--num-layers', type=int, default=3,
                     help='number of stacked RNN layers')
 parser.add_argument('--num-hidden', type=int, default=512,
@@ -57,6 +59,29 @@ parser.add_argument('--sampling', type=str, default=None,
                     help='sequence batch sampling method: random, sorted, #partitions '
                          'in sinusoidal sampling or comma separated list of buckets')
 
+
+class UtteranceBatch(object):
+  def __init__(self, data_names, data, label_names, label, bucket_key,
+               utt_names=None, utt_lens=0):
+    self.data = data
+    self.label = label
+    self.data_names = data_names
+    self.label_names = label_names
+    self.bucket_key = bucket_key
+    self.utt_names = utt_names
+    self.utt_lens = utt_lens
+
+  @property
+  def provide_data(self):
+    return [(n, x.shape) for n, x in zip(self.data_names, self.data)]
+
+  @property
+  def provide_label(self):
+    if len(self.label_names):
+      return [(n, x.shape) for n, x in zip(self.label_names, self.label)]
+    else:
+      return None
+
 class UtteranceIter(DataIter):
   """A data iterator for acoustic modeling of frame-wise labeled data.
   The iterator supports bucketing based on predefined bucket sizes,
@@ -92,7 +117,9 @@ class UtteranceIter(DataIter):
 
       self.data = [[] for _ in sampling]
       self.labels = [[] for _ in sampling]
-      for utt, lab in zip(utterances, states):
+      self.names = [[] for _ in sampling]
+      self.lengths = [[] for _ in sampling]
+      for utt, lab, name in zip(utterances, states, names):
         buck = bisect.bisect_left(sampling, len(utt))
         xin = np.full((sampling[buck],len(utt[0])), -1, dtype='float32')
         n_in = len(utt[0])
@@ -101,7 +128,8 @@ class UtteranceIter(DataIter):
         yout[:len(utt)] = lab
[…]

### 2017-04-20 code/run.py
utterance name fix

@@ -147,7 +147,6 @@ class UtteranceIter(DataIter):
     self.label_name = label_name
     self.sampling = sampling
     self.default_key = max(sampling)
-    self.names = names
     self.shuffle = shuffle
 
     # we assume time major layout
@@ -411,12 +410,12 @@ def test(args):
       outputs = model.get_outputs()
       pcx = np.log(outputs[0].asnumpy().astype('float32'))
 
-      print pcx.shape
       batch_size = len(batch.utt_names)
       pcx = pcx.reshape((pcx.shape[0]/batch_size,batch_size,pcx.shape[1]))
       pcx = pcx.swapaxes(0, 1)
 
       for i in xrange(batch_size):
+        print batch.utt_names[i]
         length = batch.utt_lens[i]
         times = zip(range(0, length), range(1, length + 1))
         cache.addFeatureCache(batch.utt_names[i] + '/1', pcx[i][:length], times)

### 2017-04-20 code/run.py
multi-batch forwarding

@@ -132,6 +132,9 @@ class UtteranceIter(DataIter):
         self.lengths[buck].append(len(utt))
       for i, buck in enumerate(self.data):
         self.idx.extend([(i, j) for j in range(0, len(buck) - batch_size + 1, batch_size)])
+        if len(buck) % batch_size != 0:
+          self.idx.append((i,len(buck)-len(buck)%batch_size)) #[(i, j) for j in range(0, len(buck) - batch_size + 1, batch_size)])
+
     else:
       raise NotImplementedError('sampling %s not supported' % str(sampling))
 
@@ -180,10 +183,13 @@ class UtteranceIter(DataIter):
     if isinstance(self.sampling, list):
       i, j = self.idx[self.curr_idx]
 
-      data = self.nddata[i][j:j + self.batch_size]
-      label = self.ndlabel[i][j:j + self.batch_size]
-      names = self.names[i][j:j + self.batch_size]
-      lens = self.lengths[i][j:j + self.batch_size]
+      slice_end = min(self.nddata[i].shape[0],j+self.batch_size)
+      #slice_end = j + self.batch_size
+
+      data = self.nddata[i][j:slice_end]
+      label = self.ndlabel[i][j:slice_end]
+      names = self.names[i][j:slice_end]
+      lens = self.lengths[i][j:slice_end]
       data = ndarray.swapaxes(data, 0, 1) # TBD
       label = ndarray.swapaxes(label, 0, 1)
 
@@ -377,7 +383,7 @@ def test(args):
         label = mx.sym.Reshape(label, shape=(-1,))
         out = mx.sym.SoftmaxOutput(data=pred, label=label, name='softmax', ignore_label=-1, multi_output=True,
                                     use_ignore=True)
-        lg = mx.sym.log_softmax(data=pred, name='softmax') #+ label
+        #lg = mx.sym.log_softmax(data=pred, name='softmax') #+ label
 
         return out, ('data',), ('labels',)
 

### 2017-04-20 code/run.py
multi-batch forwarding fix

@@ -382,8 +382,7 @@ def test(args):
 
         label = mx.sym.Reshape(label, shape=(-1,))
         out = mx.sym.SoftmaxOutput(data=pred, label=label, name='softmax', ignore_label=-1, multi_output=True,
-                                    use_ignore=True)
-        #lg = mx.sym.log_softmax(data=pred, name='softmax') #+ label
+                                   use_ignore=True)
 
         return out, ('data',), ('labels',)
 

### 2017-05-05 code/run.py
mxnet batching (wip)

@@ -72,23 +72,22 @@ class UtteranceIter(DataIter):
       either 'sorted' or 'random' 
       or number of bins in sinusoidal sampling 
       or size of data buckets (automatically generated if None).
-  layout : str
-      format of data and label. 'NT' means (batch_size, length)
-      and 'TN' means (length, batch_size).
   """
 
-  def __init__(self, utterances, states, names, batch_size, sampling, data_name='data', label_name='labels', shuffle=True):
+  def __init__(self, utterances, states, names, batch_size, batching, data_name='data', label_name='labels', shuffle=True):
     super(UtteranceIter, self).__init__()
-    if not sampling:
-      minpad = 100
-      sampling = range(minpad,max([len(s) for s in utterances]),minpad)
+    minpad = 100
+    sampling = range(minpad,max([len(s) for s in utterances]),minpad)
+    self.max_len = max([len(utt) for utt in utterances])
+    if sampling[-1] < self.max_len:
+        sampling.append(self.max_len)
 
     self.idx = []
-    if isinstance(sampling, list):
-      sampling.sort()
-      self.max_len = max([len(utt) for utt in utterances])
-      if sampling[-1] < self.max_len:
-        sampling.append(self.max_len)
+    if isinstance(batching, list):
+      batching.sort()
+      if batching[-1] < self.max_len:
+        batching.append(self.max_len)
+      sampling = batching
 
       self.data = [[] for _ in sampling]
       self.labels = [[] for _ in sampling]
@@ -105,7 +104,7 @@ class UtteranceIter(DataIter):
       for i, buck in enumerate(self.data):
         self.idx.extend([(i, j) for j in range(0, len(buck) - batch_size + 1, batch_size)])
     else:
-      raise NotImplementedError('sampling %s not supported' % str(sampling))
+      self.idx.extend([[0,j] for j in range(0, len(utterances) - batch_size + 1, batch_size)])
 
     self.data = [np.asarray(i, dtype='float32') for i in self.data] # BTD
     self.labels = [np.asarray(i, dtype='float32') for i in self.labels] # BT
@@ -113,6 +112,7 @@ class UtteranceIter(DataIter):
 
     self.batch_size = batch_size
     self.sampling = sampling
+    self.batching = batching
     self.nddata = []
     self.ndlabel = []
     self.data_name = data_name
@@ -130,8 +130,10 @@ class UtteranceIter(DataIter):
 
   def reset(self):
[…]
